---
title: "Compare result against GeneATLAS"
author: "Yuxin Zou"
date: "6/23/2019"
output: 
  html_document:
    code_folding: hide
---

We compare association results from our processed data against
GeneATLAS and Neale. Here we use phenotype height. I selected 1000 top
SNPs and 1000 null SNPs in chr1 from GeneATLAS imputed and
not-normalized (height) result.

```{r eval=FALSE}
# Run in CRI
library(readr)
library(dplyr)
# READ associations
result.file = '/gpfs/data/stephens-lab/finemap-uk-biobank/data/gene-atlas/height/imputed/not-normalized/imputed.allWhites.50-0.0.chr1.csv.gz'
result        <- suppressMessages(read_delim(result.file, delim=' '))
colnames(result) = c('SNP', 'ALLELE','iscores', 'NBETA', 'NSE', 'PV')

# Select n most strongly associated SNPs, and n random SNPs
set.seed(201906)
n = 1000
result.top = result %>% top_n(-n, PV)
result.null = result %>% filter(!(SNP %in% result.top$SNP)) %>% top_n(n, PV)

result.choose = rbind(result.top, result.null)

result.final = result.choose %>% distinct()

saveRDS(result.final, '~/imputed.allWhites.chr1.2000.SN.rds')
write.table(result.final[,1], '~/imputed.allWhites.chr1.2000.SN.txt', row.names = FALSE, col.names = FALSE, quote = FALSE)
```

Using PLINK2, we get the genotypes from UK BioBank data. This step needs about 5 hrs to finish.
```
zcat /gpfs/data/stephens-lab/finemap-uk-biobank/height.csv.gz | awk 'BEGIN{FS=","} {print $1 OFS $1}' | tail -n +2 > height.id.txt

plink2 --sample /gpfs/data/xhe-lab/uk-biobank/data/genotypes/ukb27386_imp_v3_s487324.sample \
--bgen /gpfs/data/pierce-lab/uk-biobank-genotypes/ukb_imp_chr1_v3.bgen \
--extract imputed.allWhites.chr1.2000.SN.txt \
--keep height.id.txt \
--maf 0.01 \
--make-pgen --threads 8 --memory 38000 \
--out height.chr1.imputed.2000.SN > plink2.imputed.log

plink2 --pfile height.chr1.imputed.2000.SN --threads 8 --export A-transpose --out height.chr1.imputed.2000.SN

gzip height.chr1.imputed.2000.SN.traw
```

We filtered the SNPs using MAF 1%. There are 1173 SNPs and 274549 individuals.

We performed linear model with the following covariates:

1. sex
2. age and $age^2$
3. assessment_centre
4. genotype_measurement_batch
5. principal components 1–20

```{r}
res = read.delim('../data/height.chr1.imputed.2000.plink.height.glm.linear')
geneatlas = readRDS('../data/imputed.allWhites.chr1.2000.SN.rds')
```

```{r eval=FALSE}
# Neeale data
neale.file = '/gpfs/data/stephens-lab/finemap-uk-biobank/data/summary/neale/50_irnt.gwas.imputed_v3.both_sexes.tsv.gz'
neale       <- suppressMessages(read_delim(neale.file, delim='\t'))
neale.pos = do.call(rbind, strsplit(neale$variant, "\\:"))
snps.pos = which(neale.pos[,2] %in% res$POS)
snps.chr = which(neale.pos[,1] == "1")
snps = intersect(snps.pos, snps.chr)
snps = match(neale.pos[snps,2], res$POS)
snps.pos = as.numeric(neale.pos[snps,2])
neale.choose = neale[snps,]
neale.choose$POS = snps.pos
# 1170 rows
saveRDS(neale.choose, '~/imputed.allWhites.chr1.2000.neale.SN.rds')
```

```{r}
neale = readRDS('../data/imputed.allWhites.chr1.2000.neale.SN.rds')
```

## Top SNPs

Comparing p values of top SNPs, the p values from our processed data
tend to be larger.

### GeneATLAS
```{r}
ids   <- intersect(res$ID,geneatlas$SNP[1:1000])
rows1 <- match(ids,res$ID)
rows2 <- match(ids,geneatlas$SNP)
res.strong = res[rows1,]
par(mfrow=c(1,2))
{plot(geneatlas$PV[rows2], res.strong$P, xlab = 'GeneAtlas', ylab = 'result', main='p values')
abline(0,1)}
{par(pty='s')
plot(-log10(geneatlas$PV[rows2]), -log10(res.strong$P), xlab = 'GeneAtlas', ylab = 'result', main='-log10(p values)', xlim=c(20,150), ylim=c(20,150))
abline(0,1)}
```

Plot of t statistics:
```{r}
par(pty='s')
plot(geneatlas$NBETA[rows2]/geneatlas$NSE[rows2], res.strong$T_STAT, xlab='GeneAtlas', ylab = 'result', main = 't statistics', xlim=c(-30,30), ylim=c(-30,30))
abline(0,-1)
```

### Neale

Comparing with Neale's result
```{r}
par(mfrow=c(1,2))
POS <- intersect(res.strong$POS, neale$POS)
rows1 <- match(POS,res.strong$POS)
rows2 <- match(POS,neale$POS)
neale.strong = neale[rows2,]
{plot(neale.strong$pval, res.strong$P[rows1], xlab = 'Neale', ylab = 'result', main='p values')
abline(0,1)}
{par(pty='s')
plot(-log10(neale.strong$pval), -log10(res.strong$P[rows1]), xlab = 'Neale', ylab = 'result', main='-log10(p values)', xlim=c(20,110), ylim=c(20,110))
abline(0,1)}
```

Plot of t statistics:
```{r}
par(pty='s')
plot(neale.strong$tstat, res.strong$T_STAT[rows1], xlab='Neale', ylab = 'result', main = 't statistics', xlim=c(-20,20), ylim = c(-20,20))
abline(0,-1)
```

## Null SNPs

### GeneATLAS

Comparing p values of null SNPs, the p values vary a lot!
```{r}
ids   <- intersect(res$ID,geneatlas$SNP[1001:2000])
rows1 <- match(ids,res$ID)
rows2 <- match(ids,geneatlas$SNP)
res.null = res[rows1,]
par(mfrow=c(1,2))
{plot(geneatlas$PV[rows2], res.null$P, xlab = 'GeneAtlas', ylab = 'result', main='p values')
abline(0,1)}
{plot(-log10(geneatlas$PV[rows2]), -log10(res.null$P), xlab = 'GeneAtlas', ylab = 'result', main='-log10(p values)')
abline(0,1)}
```

Plot of t statistics:
```{r}
par(pty='s')
plot(geneatlas$NBETA[rows2]/geneatlas$NSE[rows2], res.null$T_STAT, xlab='GeneAtlas', ylab = 'result', main='t statistics', xlim=c(-2.5,2.5), ylim=c(-2.5,2.5))
abline(0,1)
```

### Neale

Comparing with Neale's result
```{r}
par(mfrow=c(1,2), pty='s')
POS <- intersect(res.null$POS, neale$POS)
rows1 <- match(POS,res.null$POS)
rows2 <- match(POS,neale$POS)
neale.null = neale[rows2,]
{plot(neale.null$pval, res.null$P[rows1], xlab = 'Neale', ylab = 'result', main='p values')
abline(0,1)}
{plot(-log10(neale.null$pval), -log10(res.null$P[rows1]), xlab = 'Neale', ylab = 'result', main='-log10(p values)', xlim=c(0,2), ylim=c(0,2))
abline(0,1)}
```

Plot of t statistics:
```{r}
par(pty='s')
plot(neale.null$tstat, res.null$T_STAT[rows1], xlab='Neale', ylab = 'result', main = 't statistics', xlim=c(-2,2), ylim=c(-2,2))
abline(0,-1)
```
