---
title: "Compare result against GeneATLAS"
author: "Yuxin Zou"
date: "6/14/2019"
output: 
  html_document:
    code_folding: hide
---

We compare association results from our processed data against GeneATLAS. Here we use phenotype height. I selected 1000 top SNPs and 1000 random SNPs in chr1 from GeneATLAS imputed and not-normalized (height) result.

```{r eval=FALSE}
# Run in CRI
library(readr)
library(dplyr)
# READ associations
result.file = '/gpfs/data/stephens-lab/finemap-uk-biobank/data/gene-atlas/height/imputed/not-normalized/imputed.allWhites.50-0.0.chr1.csv.gz'
result        <- suppressMessages(read_delim(result.file, delim=' '))
colnames(result) = c('SNP', 'ALLELE','iscores', 'NBETA', 'NSE', 'PV')

# Select n most strongly associated SNPs, and n random SNPs
set.seed(201906)
n = 1000
result.top = result %>% top_n(-n, PV)
result.random = result %>% filter(!(SNP %in% result.top$SNP)) %>% sample_n(n, replace = F)

result.choose = rbind(result.top, result.random)

result.final = result.choose %>% distinct()

saveRDS(result.final, '~/imputed.allWhites.chr1.2000.rds')
write.table(result.final[,1], '~/imputed.allWhites.chr1.2000.txt', row.names = FALSE, col.names = FALSE, quote = FALSE)
```

Using PLINK2, we get the genotypes from UK BioBank data. This step needs about 5 hrs to finish.
```
zcat /gpfs/data/stephens-lab/finemap-uk-biobank/height.csv.gz | awk 'BEGIN{FS=","} {print $1 OFS $1}' | tail -n +2 > height.id.txt

plink2 --sample /gpfs/data/xhe-lab/uk-biobank/data/genotypes/ukb27386_imp_v3_s487324.sample \
--bgen /gpfs/data/pierce-lab/uk-biobank-genotypes/ukb_imp_chr1_v3.bgen \
--extract imputed.allWhites.chr1.2000.txt \
--keep height.id.txt \
--maf 0.01 \
--make-pgen --threads 8 --memory 38000 \
--out height.chr1.imputed.2000 > plink2.imputed.log

plink2 --pfile height.chr1.imputed.2000 --threads 8 --export A-transpose --out height.chr1.imputed.2000

gzip height.chr1.imputed.2000.traw
```

We filtered the SNPs using MAF 1%. There are 1211 SNPs and 274549 individuals.

We performed linear model with the following covariates:

1. sex
2. age and $age^2$
3. assessment_centre
4. genotype_measurement_batch
5. principal components 1–20

```{r}
res = readRDS('../data/height.chr1.imputed.2000.result.rds')
geneatlas = readRDS('../data/imputed.allWhites.chr1.2000.rds')
```

## Top SNPs

Comparing p values of top SNPs, the p values from our processed data tend to be larger.
```{r}
ids   <- intersect(rownames(res),geneatlas$SNP[1:1000])
rows1 <- match(ids,rownames(res))
rows2 <- match(ids,geneatlas$SNP)
par(mfrow=c(1,2))
{plot(geneatlas$PV[rows2], res[rows1,'pv'], xlab = 'GeneAtlas', ylab = 'result', main='p values')
abline(0,1)}
{plot(-log10(geneatlas$PV[rows2]), -log10(res[rows1,'pv']), xlab = 'GeneAtlas', ylab = 'result', main='-log10(p values)')
abline(0,1)}
```

Plot of estimated coefficients:
```{r}
plot(geneatlas$NBETA[rows2], res[rows1,'beta'], xlab='GeneAtlas', ylab = 'result', main = 'beta')
abline(0,1)
```

## Random SNPs

Comparing p values of random SNPs, the p values vary a lot!
```{r}
ids   <- intersect(rownames(res),geneatlas$SNP[1001:2000])
rows1 <- match(ids,rownames(res))
rows2 <- match(ids,geneatlas$SNP)
par(mfrow=c(1,2))
{plot(geneatlas$PV[rows2], res[rows1,'pv'], xlab = 'GeneAtlas', ylab = 'result', main='p values')
abline(0,1)}
{plot(-log10(geneatlas$PV[rows2]), -log10(res[rows1,'pv']), xlab = 'GeneAtlas', ylab = 'result', main='-log10(p values)')
abline(0,1)}

```

```{r}
plot(geneatlas$NBETA[rows2], res[rows1,'beta'], xlab='GeneAtlas', ylab = 'result', main='beta')
abline(0,1)
```
